---
title: "US State-Level Data Maps"
---

This page presents choropleth maps using data from the `poliscidata` package. I chose two variables: **carfatal07** (numeric) and **cook_index3** (categorical).

**Data source:** `poliscidata::states`



## Setup & Data Preparation

```{r}
#| echo: true
#| message: false
#| warning: false

# Load libraries
library(tidyverse)
library(maps)
library(sf)
library(leaflet)
library(viridis)
library(poliscidata)

# 1. Get geographic data for static maps (maps package)
states_polygon <- as_tibble(map_data("state")) |>
  select(region, group, order, lat, long)

# 2. Get geographic data for interactive maps (sf package)
states_sf <- read_sf("https://rstudio.github.io/leaflet/json/us-states.geojson") |>
  select(name, geometry)

# 3. Load poliscidata and prepare variables
polisci_data <- as_tibble(poliscidata::states) |>
  select(state, carfatal07, cook_index3)

# Clean keys for merging: lower case and remove spaces
polisci_data <- polisci_data |>
  mutate(state = str_to_lower(str_squish(state)))

states_polygon <- states_polygon |>
  mutate(region = str_to_lower(region))

states_sf <- states_sf |>
  mutate(name = str_to_lower(name))

# Merge for static plot
static_data <- states_polygon |>
  left_join(polisci_data, by = c("region" = "state")) |>
  filter(!region %in% c("alaska", "hawaii"))

# Merge for interactive plot
sf_data <- states_sf |>
  left_join(polisci_data, by = c("name" = "state")) |>
  filter(!name %in% c("alaska", "hawaii", "puerto rico")) |>
  mutate(geometry = st_make_valid(geometry))
```

### Numeric Variable: Traffic Fatalities (carfatal07)
Car crash fatalities per 100,000 people in 2007.
#### Static Map (ggplot2)
```{r}
#| echo: true
#| fig-height: 6
#| fig-width: 8

static_data |>
  ggplot(aes(x = long, y = lat, group = group)) + 
  geom_polygon(aes(fill = carfatal07), color = "black", linewidth = 0.2) + 
  coord_map() + 
  theme_void() + 
  scale_fill_viridis_c(name = "Fatalities\n(per 100k)",
                       option = "plasma") +
  labs(title = "Vehicle Crash Fatalities by State (2007)")
```

#### Interactive Map (leaflet)

```{r}
#| echo: true

# Create color palette
pal_num <- colorNumeric("viridis", domain = sf_data$carfatal07)

# Create interactive labels
sf_data <- sf_data |>
  mutate(labels_num = str_c(str_to_title(name), ": ", 
                            round(carfatal07, 1), " per 100k"))

leaflet(sf_data) |>
  addTiles() |>
  setView(-96, 37.8, 4) |>
  addPolygons(
    fillColor = ~pal_num(carfatal07),
    weight = 1,
    fillOpacity = 0.7,
    color = "white",
    label = ~labels_num,
    labelOptions = labelOptions(textsize = "12px"),
    highlightOptions = highlightOptions(weight = 3, color = "#666")
  )
```

#### Insights
The numeric map reveals significant regional variation. Southern states like Mississippi and West Virginia show the highest fatality rates, while states in the Northeast generally have lower rates. This pattern may reflect differences in rural versus urban driving conditions, speed limits, and emergency response capabilities.

### Categorical Variable: Cook Partisan Voting Index (cook_index3)
*Cook Political Report partisan lean: "D+8+", "Competitive", or "R+8+".*

#### Static Map (ggplot2)
```{r}
#| echo: true
#| fig-height: 6
#| fig-width: 8

static_data |>
  mutate(cook_cat = factor(cook_index3, 
                          levels = c("D+8+", "Competitive", "R+8+"))) |>
  ggplot(aes(x = long, y = lat, group = group)) + 
  geom_polygon(aes(fill = cook_cat), color = "black", linewidth = 0.2) + 
  coord_map() + 
  theme_void() + 
  scale_fill_manual(name = "Partisan Lean",
                    values = c("D+8+" = "blue", 
                               "Competitive" = "purple", 
                               "R+8+" = "red"),
                    na.value = "grey90") +
  labs(title = "Cook Partisan Voting Index by State")
```

#### Interactive Map (leaflet)

```{r}
#| echo: true

# Prepare data for interactive map
sf_data <- sf_data |>
  mutate(cook_cat = factor(cook_index3, 
                          levels = c("D+8+", "Competitive", "R+8+")))

# Create color palette for categorical variable
pal_cat <- colorFactor(c("blue", "purple", "red"), 
                       domain = sf_data$cook_cat,
                       na.color = "grey90")

# Create labels
sf_data <- sf_data |>
  mutate(labels_cat = str_c(str_to_title(name), ": ", cook_index3))

leaflet(sf_data) |>
  addTiles() |>
  setView(-96, 37.8, 4) |>
  addPolygons(
    fillColor = ~pal_cat(cook_cat),
    weight = 1,
    fillOpacity = 0.7,
    color = "white",
    label = ~labels_cat,
    labelOptions = labelOptions(textsize = "12px"),
    highlightOptions = highlightOptions(weight = 3, color = "#666")
  ) |>
  addLegend(pal = pal_cat, values = ~cook_cat, opacity = 0.7,
            title = "Partisan Lean", position = "bottomright")
```

#### Insights
The categorical map shows a clear geographic pattern of partisanship. Democratic-leaning states (blue) cluster in the Northeast and West Coast, while Republican-leaning states (red) dominate the South and Midwest. Competitive states (purple) are scattered, often around the Great Lakes region.

## Summary
This page demonstrates four choropleth maps (static + interactive for both numeric and categorical variables) using real US state data from the poliscidata package. Interactive maps allow users to hover over each state to see exact values.
